---
title: "Introduction to glemb"
author: "Paul D. Allison"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to glemb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`glemb` performs multiple imputation of data sets containing a mix of
continuous and categorical variables. It implements the General Location
Model (GLM) of Olkin and Tate (1961) and Little and Schluchter (1987),
combined with the Expectation-Maximization with Bootstrapping (EMB)
algorithm of Honaker, King, and Blackwell (2011).

Readers familiar with multiple imputation will recognize two common
approaches: Markov chain Monte Carlo (MCMC), as implemented in Schafer's
`mix` package, and fully conditional specification (FCS), as implemented
in `mice`. The `glemb` package offers a third approach based on a principled
joint model for mixed data. Compared with MCMC, it avoids the need to monitor
convergence of a Markov chain. Compared with FCS, it guarantees that the
imputation model corresponds to a proper joint distribution.

The General Location Model factorizes the joint distribution of continuous
variables **y** and categorical variables **x** as:

$$p(\mathbf{y}, \mathbf{x}) = p(\mathbf{y} \mid \mathbf{x})\, p(\mathbf{x})$$

The categorical margin $p(\mathbf{x})$ is modelled with a log-linear model.
The continuous conditional $p(\mathbf{y} \mid \mathbf{x} = c)$ is
multivariate normal with a cell-specific mean and a covariance matrix shared
across all cells. This implies that the conditional mean of every continuous
variable is a linear function of all the other variables (both continuous and
categorical), and that the conditional distribution of every categorical
variable is a logistic regression function of all the other variables.

Proper multiple imputations are generated via the EMB algorithm: for each of
the `m` imputations, a bootstrap sample is drawn from the incomplete data,
the ECM algorithm (a variant of EM) is run on the bootstrap sample to obtain
maximum likelihood estimates of the parameters, and one completed data set
is drawn from the posterior predictive distribution using those parameters.

## Example: Antisocial Behavior in the NLSY

We illustrate `glemb` using data from the National Longitudinal Survey of
Youth (NLSY). The data set contains 581 children with measurements on seven
variables:

| Variable  | Description |
|-----------|-------------|
| `anti`    | Antisocial behavior score, ranging from 0 to 6 |
| `self`    | Self-esteem score, ranging from 6 to 24 (148 missing) |
| `pov`     | Poverty status: 1 = in poverty, 0 = otherwise (150 missing) |
| `race`    | Race/ethnicity: 1 = White non-Hispanic, 2 = Black, 3 = Hispanic (113 missing) |
| `momwork` | Mother employed in 1990: 1 = yes, 0 = no (86 missing) |
| `divorce` | Mother ever divorced: 1 = yes, 0 = no |
| `gender`  | Child's gender: 1 = female, 0 = male |

The goal is to regress `anti` on all other variables after imputing the
missing values in `self`, `pov`, `race`, and `momwork`.

## Running glemb

After loading the data, we select the variables of interest and call
`glemb()`. The `noms` argument identifies the categorical variables; all
remaining variables are treated as continuous.

```{r impute, eval = FALSE}
library(glemb)
library(rio)
library(dplyr)
library(mice)
library(knitr)

nlsy <- rio::import("https://statisticalhorizons.com/wp-content/uploads/2022/01/nlsymiss.dta")

nlsy <- nlsy %>%
  select(pov, race, momwork, divorce, gender, self, anti)

imp <- glemb(nlsy, m = 50,
             noms = c(pov, race, momwork, divorce, gender),
             seed = 456)
```

The `noms` argument accepts variable names either quoted (`"pov"`) or
unquoted (`pov`). All variables listed in `noms` are returned as factors
in the completed data sets, ensuring they are treated as categorical in
downstream regression calls.

By default, `glemb()` returns a `mids` object (from the `mice` package),
which can be passed directly to `with()` and `mice::pool()` for analysis
and pooling. Optionally, `glemb` can return a plain list containing the
completed data frames by specifying `output = "list"`.

## Fitting and Pooling the Analysis Model

We fit a linear regression model to each of the 50 completed data sets and
pool the results using Rubin's rules:

```{r analyse, eval = FALSE}
fit <- with(imp, lm(anti ~ pov + race + momwork + divorce + gender + self))

kable(summary(mice::pool(fit)), digits = 3)
```

The `with()` function applies the model to each completed data set, and
`mice::pool()` combines the estimates and standard errors using Rubin's
(1987) rules. Because all `noms` variables are returned as factors,
`race` is automatically treated as a categorical variable with two dummy
variables (Black vs. White non-Hispanic and Hispanic vs. White non-Hispanic)
in the regression.

The results are shown below:

```{r results, echo = FALSE}
results <- data.frame(
  term      = c("(Intercept)", "pov1", "race2", "race3", "momwork1",
                "divorce1", "gender1", "self"),
  estimate  = c( 3.006,  0.634,  0.073, -0.352,  0.216, -0.094, -0.561, -0.068),
  std.error = c( 0.457,  0.168,  0.157,  0.161,  0.141,  0.147,  0.118,  0.021),
  statistic = c( 6.576,  3.777,  0.463, -2.182,  1.533, -0.644, -4.778, -3.238),
  df        = c(371.378, 196.289, 337.789, 431.635, 343.801, 517.182, 550.673, 371.954),
  p.value   = c( 0.000,  0.000,  0.643,  0.030,  0.126,  0.520,  0.000,  0.001)
)
knitr::kable(results, digits = 3)
```

Children in poverty score 0.634 points higher on the antisocial behavior
scale than children not in poverty, adjusting for all other variables.
Hispanic children score 0.352 points lower than White non-Hispanic children.
Girls score 0.561 points lower than boys, and higher self-esteem is associated
with lower antisocial behavior.

## References

Honaker, J., King, G., & Blackwell, M. (2011). Amelia II: A program for
missing data. *Journal of Statistical Software*, 45(7), 1--47.

Little, R. J. A., & Schluchter, M. D. (1987). Maximum likelihood estimation
for mixed continuous and categorical data with missing values.
*Biometrika*, 74, 165--173.

Olkin, I., & Tate, R. F. (1961). Multivariate correlation models with mixed
discrete and continuous variables. *Annals of Mathematical Statistics*,
32, 448--465.

Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys*.
Wiley.

Schafer, J. L. (1997). *Analysis of Incomplete Multivariate Data*.
Chapman & Hall.
